---
title: Gini Index under Fat-Tails
author: ''
date: '2020-06-26'
slug: gini-index-under-fat-tails
categories: []
tags: []
aliases: 
  - ../../2020/06/26/gini-index-under-fat-tails/
---

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(reldist)
# Roboto-Condensed
extrafont::loadfonts(device="win")
# set seed
set.seed(45642)
# set theme
theme_set(hrbrthemes::theme_ipsum_rc(grid = "Y"))
```

I have recently been exploring Nassim Taleb's latest [technical book: Statistical Consequences of Fat Tails](https://www.researchers.one/media/documents/260-m-Technical%20Incerto%20Vol%201.pdf). In this blogpost, I'll follow Taleb's exposition of the Gini Index under fat-tails in Chapter 13 of his book. 

Intuitively, if we use the "empirical distribution" to estimate the Gini Index, under fat-tails, we underestimate the tail of the distribution and thus underestimate the Gini index. This is [yet another example](https://david-salazar.github.io/2020/06/11/how-to-not-get-fooled-by-the-empirical-distribution/) of how we *fool* ourselves when we are using the "empirical" distribution. Instead, Taleb recommends first understanding the tail behavior of the "Gini Index" [by estimating the tail index of the distribution with Maxmimum Likelihood](https://david-salazar.github.io/2020/06/11/how-to-not-get-fooled-by-the-empirical-distribution/) and *then* using the **functional form** of the maximum likelihood estimator for the Gini Index. 

## What is the Gini Index?

The Gini index is a measure of concentration commonly used in the income and wealth inequalities discussions. The stochastic representation of the Gini $g$ is:

$$
g=\frac{1}{2} \frac{\mathbb{E}\left(\left|X^{\prime}-X^{\prime \prime}\right|\right)}{\mu} \in[0,1]
$$

where \( X^{\prime} \) and \( X^{\prime \prime} \) are i.i.d. copies of a random variable \( X \) with c.d.f. \( F(x) \in[c, \infty) \) \( c>0, \) and with finite mean \( \mathbb{E}(X)=\mu . \) 

Intuitively, then:

> The Gini Index of a random variable X is the mean expected deviation between any two independent realizations of X, scaled twice by the mean

## Non-Parametric estimator

Using the empirical distribution, then, we can estimate the Gini Index $g$:

$$
G^{N P}\left(X_{n}\right)=\frac{\sum_{1 \leq i<j \leq n}\left|X_{i}-X_{j}\right|}{(n-1) \sum_{i=1}^{n} X_{i}}
$$

We will examine the behavior of this estimator both under thin-tailed distributions and under fat-tailed distributions. 

### Finite variance, asymptotic normality for the non parametric estimator

Under the hypothesis of finite variance for the data-generating process of $X$, the estimator is asymptotically normal. We can check this condition with Monte-Carlo simulations. I'll perform $10^4$ Monte-Carlo experiments: in each of them, I'll generate a 1000 samples from a lognormal with underlying standard deviation of 0.2, which behaves like a Gaussian. Then, I'll calculate the Gini estimate using the non-parametric estimator. 

Given this lognormal, the "true" Gini is thus:

$$
G=2 \Phi\left(\frac{\sigma}{\sqrt{2}}\right)-1 = 0.11246
$$

```{r lognormal, fig.width=8}
crossing(experiment = 1:10^4,
         sample_size = 1000) %>% 
  mutate(data = map(sample_size, ~ rlnorm(., sdlog = 0.2)),
         gini = map_dbl(data, ~ gini(.))) -> gini_lognormal

gini_lognormal %>% 
  ggplot(aes(gini)) +
  geom_histogram(binwidth = 0.0005, color = "black", fill = "dodgerblue4", alpha = 0.7) +
  geom_vline(aes(xintercept = 0.11246), color = "red", linetype = 2) +
  labs(title = "Non parametric Gini estimator",
       subtitle = "Under finite variance, non parametric estimator is asymptotically normal",
       caption = "Data-generating process is lognormal with underlying sd = 0.2, which behaves like a Gaussian")
```           

### Fat-tails: Non-parametric estimator

However, when the data-generating process of $X$ is in [the MDA of the Fréchet](https://david-salazar.github.io/2020/06/10/fisher-tippet-th-a-clt-for-the-sample-maxima/) (i.e., it's a fat-tailed variable), the non-parametric estimator of the Gini Index loses its properties of normality. Indeed, the limiting distribution of the non-parametric index becomes a skewed-to-the-right $\alpha$-stable law. Thus, the non-parametric estimate underestimates the true Gini Index.

This can be checked with Monte-Carlo simulations. I'll perform $10^4$ Monte-Carlo experiments: in each of them, I'll generate a 1000 samples from a Pareto with $\alpha = 1.16$. Then, I'll calculate the Gini estimate using the non-parametric estimator. 

Given this Pareto, the "true" Gini is thus:

$$
g = \dfrac{1}{2\alpha -1} = 0.7575758
$$

```{r pareto, fig.width=8}
rpareto <- function(n) {
    alpha <- 1.16
   (1/runif(n)^(1/alpha)) # inverse transform sampling
}

crossing(experiment = 1:10^4,
         sample_size = 1000) %>% 
  mutate(data = map(sample_size, ~ rpareto(.)),
         gini = map_dbl(data, ~ gini(.))) -> gini_pareto

gini_pareto %>% 
  ggplot(aes(gini)) +
  geom_histogram(binwidth = 0.01, color = "black", fill = "dodgerblue4", alpha = 0.7) +
  geom_vline(aes(xintercept = 0.7575758), color = "red", linetype = 2) +
  annotate("text", x = 0.78, y = 400, label = "True gini", color = "red", 
           family = theme_get()$text[["family"]]) +
  labs(title = "Non parametric Gini estimator",
       subtitle = "Under fat-tails, non parametric estimator is skewed to the right. Thus, downward bias",
       caption = "Data generating process is Pareto with alpha = 1.16")
```

Therefore, under fat-tails, **the non-parametric Gini estimator will approach its true value more slowly, and from below.** 

## The Maximum Likelihood alternative

A better alternative when working with fat-tails, it's to first estimate the tail and then derive your quantity of interest. Indeed, one does not need too much data to derive the properties of the tail. With a Pareto, for example, the Maximum Likelihood estimator for the tail exponent follows an inverse Gamma distribution that rapidly converges to a Gaussian tightly *around* the true $\alpha$. Therefore, one can reliably estimate the tail exponent of the Pareto and thus understand the properties of the distribution with relatively few data.

The ML estimator for the tail exponent of a Pareto is thus:

$$
\widehat \alpha = \frac{n}{\sum _i  \ln (x_i) }
$$
Then, we can derive our Maximum Likelihood estimate for the Gini Index:

$$
g = \dfrac{1}{2\widehat \alpha -1}
$$

Indeed, Taleb shows that this estimator for the Gini Index is **not just asymptotically normal, but also asymptotically efficient**. We can test for these using our Monte-Carlo simulations. For each of our simulated datasets, we can derive our Maximum Likelihood estimate and then derive our Maximum Likelihood estimate for the Gini Index.

```{r comparison, fig.width=10, fig.height=7}
estimate_alpha_ml <- function(observations) {
  alpha <- length(observations)/sum(log(observations))
  if (alpha < 1) {
    alpha <- 1.0005 
  }
  alpha
}

gini_pareto %>% 
  mutate(alpha_ml = map_dbl(data, ~ estimate_alpha_ml(.)),
         gini_ml = 1/(2*alpha_ml - 1)) -> gini_pareto

gini_pareto %>%
  rename(nonparametric = gini,
         maximum_likelihood = gini_ml) %>% 
  pivot_longer(c(nonparametric, maximum_likelihood), names_to = "estimator", values_to = "gini") %>% 
  ggplot(aes(gini, fill = estimator)) +
  geom_histogram(binwidth = 0.01, color = "black", alpha = 0.7,
                 position = "identity") +
  geom_vline(aes(xintercept = 0.7575758), color = "red", linetype = 2) +
  annotate("text", x = 0.78, y = 1000, label = "True gini", color = "red", 
           family = theme_get()$text[["family"]]) +
  scale_fill_viridis_d() +
  theme(legend.position = "bottom") +
  labs(title = "Comparison of Gini estimators: Non-parametric vs Maximum Likelihood",
       subtitle = "Under fat-tails, unlike the non parametric estimator, Max Likelihood estimate is still asymptotically normal",
       caption = "Data generating process is a Pareto with alpha = 1.16")
```
## Conclusion

When the underlying distribution is fat-tailed, which is always in the case of income or wealth, the non-parametric estimator for the Gini index is skewed to the right and thus underestimates the true Gini index. In this case, it is a much statistically sound strategy to first estimate the tail behavior of the distribution with Maximum Likelihood and then estimate the Gini Index with its plug-in estimator. 
