---
title: Plotting Supply and Demand Curves with ggplot2
subtitle: ggsupplyDemand
author: David Salazar
date: '2018-05-20'
slug: ggsupplydemand
categories:
  - Packages
tags:
  - Ggplot2
---

# What is it?

`ggsupplyDemand` is an R [package](https://github.com/David-Salazar/ggsupplyDemand) that makes it extremely easy to plot basic supply and demands using ggplot2

```{r message=FALSE, warning=FALSE}
library(ggsupplyDemand)

create_supply_and_demand() %>% 
  shift_demand(outwards = TRUE) %>% 
  plot_supply_and_demand(consumer_surplus = TRUE)
```

# Why?

I needed to plot some basic supply and demand curves in R. Obviously, I thought of `ggplot2`. However, it is not that straightforward. The best resource I could find was [this](https://www.andrewheiss.com/blog/2017/09/15/create-supply-and-demand-economics-curves-with-ggplot2/) blogpost from Andrew Heiss. I recopilated most of his functions, created a simple API and put all the functions on an [package](https://github.com/David-Salazar/ggsupplyDemand).  

## How does it work?

Build step-by-step a `tibble` that represents a system of supply and demand. Use `create_supply_and_demand()` and `shift_demand()`, `shift_supply()`. Then, use the `plot_supply_and_demand()` function on the said `tibble`. 

```{r message=FALSE, warning=FALSE}
create_supply_and_demand() %>% 
  shift_demand(outwards = TRUE) %>% 
  shift_supply(outwards = FALSE) %>% 
  plot_supply_and_demand(consumer_surplus = FALSE)
```

### Customization

`ggsupplyDemand` has a default shifter for both the demand and supply. However, if you need something different, you can always change it:

```{r message=FALSE, warning=FALSE}
create_supply_and_demand() %>% 
  shift_supply() %>% 
  shift_supply(shifter = 250) %>% 
  shift_demand(outwards = FALSE, shifter = 400) %>% 
  plot_supply_and_demand()
```

