---
title: 'The Adam Smith Problem: Tidytext in R'
author: David Salazar
date: '2018-04-09'
slug: the-adam-smith-problem-tidytext-in-r
categories:
  - tutorial
tags:
  - tidytext
---

# Why?

This is a fun (for me) exercise to explore [Text Mining with R](https://www.tidytextmining.com) and make sure I can follow along. 

# What is it ?

Around the XIX century, some german scholars posited that Wealth of Nations' Adam Smith was too different to Theory of Moral Sentiments' Adam Smith, thus concluded that he must have had a change of heart somewhere along his life or that he was simply an incoherent man. 

I tend to think that the differences between the books are overstated, and, although they treat very different subjects, they fit within a greater scientific investigation. Let's use `tidytext` principles to investigate the differences between the books. However, I do not think these similarities can be found with a quantitative bend; quite the opposite, I believe this little experiment will yield the superficial differences between the books. Let's see what we find. 

## Load packages and Data

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidytext)
library(gutenbergr)
library(tidygraph)
library(igraph)
library(ggraph)
library(stringr)
library(widyr)
```

Let's use the `gutenbergr` to access the books:

```{r}
gutenberg_works(author == "Smith, Adam")
```

Revealing that project gutenber has no "Theory of the moral Sentiments" :). Let's download the text from [here](http://ota.ox.ac.uk/desc/3189).

```{r}
wealth <- gutenberg_download(gutenberg_id = 3300) %>% 
  select(-gutenberg_id) %>% 
  mutate(chapter = cumsum(str_detect(text, regex("^chapter [\\divxlc]",
                                                 ignore_case = TRUE))),
         title = "Wealth of Nations")
wealth
```

```{r}
theory <- read_file("../../data/3189.txt") %>% 
  data_frame(text = .) %>% 
    mutate(text = strsplit(text, "\n")) %>% 
    unnest(text) %>% 
    mutate(title = "The Theory of Moral Sentiments",
           chapter = cumsum(str_detect(text, regex("^Chap. [\\divxlc]",
                                                 ignore_case = TRUE))))
theory
```

# TidyText analysis

## Most used words

Let's see which are the words that Smith uses the most in each of the books. 

```{r}
smith_books <- rbind(theory, wealth)
smith_books_tidy <- smith_books %>% 
  unnest_tokens(word, text) %>%
  anti_join(stop_words)
smith_books_tidy %>% 
  count(word, title) %>% 
  group_by(title) %>% 
  arrange(desc(n)) %>% 
  top_n(10) %>% 
  ggplot(aes(x = fct_reorder(word, n), y = n, fill = title)) +
    geom_col() +
    coord_flip() +
    hrbrthemes::theme_ipsum() +
    facet_wrap(~ title, scales = "free") +
    ggthemes::scale_fill_gdocs(guide = FALSE) +
    labs(x = "")
```

Well, Smith definitely wrote two books with (just in the surface, as his theses were very similar. cf. Marketplace Smith) different subjects, as shown by the very different words he uses in both books. 

## Sentiment Analysis

Economics was famously described as the "dismal science". Let's see if the most economics-centered book is indeed much somber than the "Theory of Moral Sentiments". 

```{r message=FALSE, warning=FALSE}
smith_books_tidy %>% 
  inner_join(get_sentiments("afinn")) %>% 
  group_by(title, chapter) %>% 
  summarise(sentiment = sum(score)) %>% 
  ggplot(aes(x = chapter, y = sentiment, color = title)) +
    geom_point(alpha = 1/10) +
    geom_smooth(se = FALSE) +
    facet_wrap(~ title, scales = "free_x") +
    hrbrthemes::theme_ipsum() +
    ggthemes::scale_color_gdocs(guide = FALSE)
```

Whereas the "Theory" seems to start off neutral and end on a happy note, the "Wealth" seems to progressively get more positive but ends on a quite sad note. My former Adam Smith obsessed self may have told you why this is thus, but I forgot the chapters order and their corresponding subjects. However, there does not seem to be a ground to argue that the "Wealth of Nations" is "much dismal" than the "Theory of Moral Sentiments".

## TF-IDF

TF-IDF adjusts the term frequency within document (in this case book) for the specificity of the term to that specific document. Thus, terms that are used in both books are weighted down, and terms that are used almost exclusively in one of the books are weighted up. The idea is to find the most distinctive terms for each book. My intuition tells me that this won't be much too different from the original counts (very different vocabularies in the books). Just for the sake of it, let's use bigrams:

```{r}
smith_bigrams <- smith_books %>% 
  unnest_tokens(bigram, text, token = "ngrams", n = 2)

bigrams_separated <- smith_bigrams %>%
  separate(bigram, c("word1", "word2"), sep = " ")

bigrams_filtered <- bigrams_separated %>%
  filter(!word1 %in% stop_words$word) %>%
  filter(!word2 %in% stop_words$word) %>% 
  filter(!str_detect(word1, "\\d"),
         !str_detect(word2, "\\d"))

bigrams_united <- bigrams_filtered %>%
  unite(bigram, word1, word2, sep = " ")

bigrams_united %>% 
  count(title, bigram) %>% 
  group_by(title) %>% 
  arrange(desc(n)) %>% 
  top_n(10) %>% 
  ggplot(aes(x = fct_reorder(bigram, n), y = n, fill = title)) +
    geom_col() +
    coord_flip() +
    facet_wrap(~ title, scales = "free") +
    hrbrthemes::theme_ipsum() +
    ggthemes::scale_fill_gdocs(guide = FALSE) +
    labs(x = "")

```

With just the counts, the subject difference is even more striking. Human nature vs foreign trade. Surprsingly, one could use the bigram for the "Theory" as the main subject of the book and not be completely wrong. I do not think that the TF-IDF analysis will change much, but let's go ahead:

```{r}
bigrams_united %>%
  count(title, bigram) %>%
  bind_tf_idf(bigram, title, n) %>%
  group_by(title) %>% 
  arrange(desc(tf_idf)) %>% 
  top_n(10) %>% 
  ggplot(aes(x = fct_reorder(bigram, tf_idf), y = tf_idf, fill = title)) +
    geom_col() +
    coord_flip() +
    facet_wrap(~ title, scales = "free") +
    hrbrthemes::theme_ipsum() +
    ggthemes::scale_fill_gdocs(guide = FALSE) +
    labs(x = "")
```

As predicted, the TF-IDF adjustment didn't change much. 

## Correlating pairs

Bigrams are restrictive: words have to appear next to each other. However, words that tend to appear together, even if not immediately to each other within a given subdivision, may also give us some insight. Let's analyze which pairs of words are more likely to co-appear within a given chapter than appearing by themselves, using the \phi correlation coefficient. 

```{r}
pairwise_book <- function(df) {
  df %>% 
  group_by(word) %>%
  filter(n() >= 100) %>%
  pairwise_cor(word, chapter, sort = TRUE)
}

smith_correlations <- smith_books_tidy %>% 
  filter(!str_detect(word, "\\d")) %>% 
  group_by(title) %>% 
  do(pairwise_book(.)) %>% 
  group_by(title) %>% 
  arrange(desc(correlation)) %>% 
  top_n(50, correlation) %>% ungroup

corr_graph <- function(df) {
  ggraph(df, layout = "fr") +
  geom_edge_link(aes(edge_alpha = correlation), show.legend = FALSE) +
  geom_node_point(color = "lightblue", size = 3) +
  geom_node_text(aes(label = name), repel = TRUE) +
  theme_graph() 
}

smith_data <- smith_correlations %>%
  filter(correlation > 0.1) %>% 
  group_by(title) %>% 
  nest()

smith_data$data %>% 
  map(graph_from_data_frame) %>% 
  map(corr_graph) %>% 
  reduce(cowplot::plot_grid, labels = c("Wealth of Nations", "Theory of Moral Sentiments"))
  
```

Whereas the words from the "Theory" form a nice little cluster, the words from the "Wealth of Nations" are all over the place. 

## Conclusion

Where one to argue that Smith's two most famous book represent an incongruent body of work, a cursory look at the vocabulary would be one's best weapon. One talks about human nature and moral sentiments whereas the other takes about trade and produce. If pushed further, one could say the difference it is that whereas in the "Theory" greed is criticised, in the "Wealth" it is celebrated. However, I do believe that saying so would be a mistake. 








